---
title: 'Final Project: Major League Baseball'
author: "Kyle Sorstokke"
date: "6/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Plate Discipline and Offensive Success

In this analysis, I am seeking to understand the relationship between plate
discipline (swinging at fewer pitches outside the strike zone) and offensive
success, which I define using OPS+, a standardized measurement of on base
percentage and slugging percentage around the league average of 100. Baseball
coaches and players have historically argued over how aggressive an approach
yields the most success, but experts seem to agree that selectivity is the
most beneficial. With this in mind, my guiding question is: is plate
discipline, defined by out of zone swing %, indicative of greater offensive
success in terms of OPS+? I am also seeking to answer whether or not players
can be grouped by plate discipline to better predict OPS+ since that approach
is often used in practical baseball application. I will conduct my analysis
using data from statcast, an automatic data collection tool created by MLB in
2015. The data is sourced from baseball savant, MLB's official website for
public statcast data.


``` {r, include = FALSE}
# Reading in the statcast data
statcast <- read_csv("../data/stats.csv")

# Shuffling data for train/test splitting
set.seed(123)
statcast <- statcast[sample(nrow(statcast)),]

# root mean squared error for true y and prediction yhat
rmse <- function(y,yhat) sqrt(mean((y-yhat)^2))
```

``` {r}
# Mutating statcast dataframe to include OPS+ and groupings of plate approach
statcast <- statcast %>%
  mutate(opsPlus = (on_base_percent / mean(on_base_percent) +
                      slg_percent / mean(slg_percent) - 1) * 100) %>%
  mutate(ozSwingGroup = cut(oz_swing_percent, c(10, 20, 30, 40, 50)))

# Creating train and test data sets (90% and 10%)
statTrain <- statcast[1:1592,]
statTest <- statcast[1593:nrow(statcast),]
```

``` {r, echo = FALSE}
hist(statcast$oz_swing_percent,
     main = "Distribution of Out of Zone Swing Percent",
     xlab = "Out of zone swing %",
     ylab = "Number of Player-Seasons")

hist(statcast$opsPlus,
     main = "Distribution of OPS+",
     xlab = "OPS+",
     ylab = "Number of Player-Seasons")
```

The histogram of OPS+ proves that the average player has an OPS+ of 100,
with the majority of players clustered around the mean in a fairly normal
distribution. This distribution is the point of OPS+; it takes OPS (on-base
plus slugging), an unstandardized continuous metric for offensive output, and
modifies it in relation to the league average.

``` {r, echo = FALSE}
ggplot(statcast, aes(ozSwingGroup, opsPlus)) +
  geom_hline(yintercept = 100, color = "red", size = 2) +
  geom_boxplot(varwidth=T, fill="skyblue3") + 
  labs(title="Batter Success by Plate Discipline Groups",
       x="Out of Zone Swing Percent",
       y="OPS+")
```

``` {r}
opsMod <- aov(opsPlus ~ ozSwingGroup, data = statTrain)
summary(opsMod)
paste("RMSE:", rmse(statTest$opsPlus, predict(opsMod, statTest)))
```

I selected a one-way ANOVA test to understand how players can be grouped by
out of zone swing % to analyze their resulting offensive success. For the
outcome I selected OPS+ because it is one of the most respected metrics of
individual offensive production, standardizing On-base plus slugging to a
league average of 100, with a player at 150 producing 50% more offensive
value. OPS+ typically only compares players from a single season, but I chose
to use the entire statcast era (since data collection started in 2015) to
increase the sample size and get a better understanding of overall trends.
While out of zone swing % is a continuous variable, I grouped
it as a factor because MLB analysts often seek to understand different
approaches to hitting; one major factor in plate approach is discipline
(swinging at good pitches in the strike zone and not pitches outside the
strike zone). With this in mind, I created groups from out of zone swing %
to create groups of players with similar plate discipline. The one-way ANOVA
test allows me to determine if the groups do in fact have different sample
means, which would prove that plate discipline does effect on offensive success. 

In a one-way ANOVA test, the F value represents the ratio of variance between
samples to variance within samples; the higher the value, the greater the
relative difference in sample means. The common rejection level says that
an F value < 1 means there is no significant difference between sample groups.
My test yields an F value of 11.91 with a p-value of essentially 0; plate
discipline, defined by out of zone swing rate, is indicative of greater
success at the plate in terms of OPS+.

``` {r, echo = FALSE}
ggplot(statcast, aes(oz_swing_percent, opsPlus)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Offensive Success by Out of Zone Swing Rate",
       x = "Out of zone swing %",
       y = "OPS+")
```

``` {r}
linMod <- lm(opsPlus ~ oz_swing_percent, data = statTrain)
summary(linMod)
paste("RMSE:", rmse(statTest$opsPlus, predict(linMod, statTest)))
```

To further analyze the relationship between out of zone swing % and OPS+,
I am using a standard linear model. The linear model has t scores of
essentially 0 for the intercept and coefficient, meaning the association
is statistically significant. Similar to the ANOVA model, the negative
coefficient on the covariate shows that the association is negative; an
increase in out of zone swing % is associated with an expected decrease in
OPS+ (-0.65 points per oz swing %). This relationship is plausible because
intuition tells us that players with more disciplined approaches at the plate
(lower oz swing %) swing at fewer bad pitches, and are subsequently more
successful hitters. 

Both models yield predictions with fairly low root mean squared errors, but
the ANOVA model has slightly better predictive power (rmse 22.79 compared to
22.98). Both models are useful representations of the association between out
of zone swings and OPS+. The slight difference in rmse tells us that the
grouping of players by their plate approach yields a slight increase in
predictive power; this indicates that the common practice of analyzing
plate approach types (groups) is statistically useful. This analyisis has
practical application potential because it can help managers predict
players' offensive success and decide who to play more.
